Program Optimisations via Hylomorphisms for Extraction of Executable Code

### 2025-06-16 paper/paper.tex
Fixing typos in Section 2

Suppose, for example, that we want to write a program that sorts a list of
integers and multiplies them by 2 at the same time. In OCaml, we may write this
function directly:
\begin{minted}{ocaml}
let rec sort_times_two = function
  | [] -> []
  | h :: t -> let (l, r) = partition (fun x -> x < h) t in
  sort_times_two l @ (h * 2) :: sort_times_two t
\end{minted}

@@

Given two functors $F, G : \CatC \to \CatD$, a \emph{natural transformation} is
a map between functors is a family of morphisms $\phi_{X} : FX \to GX$ indexed
by the objects $X \in \Obj(\CatC)$ and such that it is \emph{natural} in $X$,
that is $ G(f) \cdot \phi_{X} = \phi_{Y} \cdot F(f)$. Intuitively, a natural
transformation transforms the structure of a functor into the structure of
another functor without assuming what is the type of data contained in them, for
example, a program $f : \texttt{List }X \to \texttt{Maybe} X$ which returns
nothing if the list is empty and the head of the list otherwise needs not to
know the type of the data.

@@

For example, for a set $A$ we define the functor $F : \Set \to \Set$ mapping
$X \mapsto 1 + A \times X$. An $F$-algebra is a set $B$ together with a
structure map $[b, i] : 1 + A \times B \to B$. The initial $F$-algebra is
clearly the set of lists $\List{A}$ and the catamorphism associated with the
type of lists is the unique arrow which recursively translates the initial
algebra $[\nil,\cons]$ into the algebra $[base, step]$ while turning the operation
$\nil$ into $base$ and $\cons$ into $step$.  In functional programming this is
commonly referred to as $\foldr : (1 \to B) \to (A \times B \to B) \to \List{A} \to B$. We
can in fact set $\foldr\;base\;step = \cata{[base, step]}$.

@@

For example, for a set of states $X$ and a finite set of labels $L$ we can
define a labelled transition system (LTS) on $X$ as a function $X \to BX$
implementing the \emph{transition system} with $BX = L \times X$.  In
particular, for a state $x_{1} \in X$, $c(x_{1})$ returns a pair $(l, x_{2})$
where $l \in L$ is the observable action and $x_{2} \in X$ is the next state.
In this case, we can even set some notation for the transition map:
\[
  x \xrightarrow{l} y \eqdef c_{X}(x) = (l, y)
\]
The category of $B$-coalgebras, denoted $B\Coalg$, is the category where objects
are $B$-coalgebras and morphisms $f : (X, c_{X}) \to (Y, c_{Y})$ are
$B$-coalgebra homomorphisms $f : A \to B$, that is
$ c_{Y} \circ f = F(f) \circ c_{X}$. The terminal object in this category is
called the terminal, or final, $B$-coalgebra. The carrier of this coalgebra
corresponds to the greatest fixed-point for the functor $B$, denoted by
$(\nu B, \OutIso)$ with $\OutIso$ being the final $B$-coalgebra witnessing the
isomorphism and $\OutIso : \nu B \to B\nu B$ being its inverse.

For example, the terminal coalgebra for the functor $BX = L \times X$ is the set
of infinite streams, that is the greatest solution to the domain equation
$\Str{A} \iso A \times \Str{A}$.  The uniqueness property of terminal
$B$-coalgebras states that for any other $B$-coalgebra $(X, c_{X})$ there exists
a unique $B$-coalgebra homomorphism into the terminal coalgebra
$(\nu B, \OutIso)$ which is denoted by $\ana{c_{X}}$ and pronounced
``\emph{anamorphism}'' or ``\emph{unfold}''. We spell out the uniqueness
property:
\begin{equation}
  \label{eq:ana:unique}
  f = \ana{c_{X}} \iff f = \OutOp \circ B(f) \circ c_{X}
\end{equation}
From the uniqueness property we can derive the fusion law for unfolds:
\begin{equation}
  \label{eq:ana:fusion}
  \ana{c_{Y}} \circ f = \ana{c_{X}}
\end{equation}
for all $B$-coalgebra homomorphisms $f : (X, c_{X}) \to (Y, c_{Y})$.

@@

An example of a recursive coalgebra is the \coq{partition} function in quicksort
which destructures a list into a pivot and two sub-list and as long as the
sub-lists are smaller the partitioning function still yields a unique solution to
the recursion scheme.  The uniqueness property of the hylomorphisms yields the following
fusion laws:
\begin{align}
  \label{eq:hylo:fusion}
  f \circ \hylo{c}{a} & = \hylo{c}{a'}
  \hspace{1cm}
  \Longleftarrow
  \hspace{1cm}
  f \circ a = a' \circ F(f) \\
  \hylo{c}{a} \circ f & = \hylo{c'}{a}
  \hspace{1cm}
  \Longleftarrow
  \hspace{1cm}
  c \circ f = F(f) \circ c'
\end{align}
Using the hylomorphism fusion laws, we can prove the well-known \emph{deforestation
optimisation}, also known as the \emph{composition
law}~\cite{DBLP:conf/ifl/HinzeHJ10}. This is when two consecutive recursive
computations, one that builds a data structure, and another one that consumes
it, can be fused together into a single recursive definition. This, in turn,
allows us to prove that a recursive hylomorphism is the composition of a
catamorphism and a recursive anamorphism.

@@

\section{Formalising Recursion Schemes}
\label{sec:recursion-schemes}
%Recursion schemes provide an abstract way to consume and generate data.
We now consider the formalisation hylomorphisms in Coq. We first formalise
container functors as a tool to represent polynomial functors
(Section~\ref{sec:containers}). Then we formalise algebras for container
extensions (Section~\ref{sec:coq-algebras}), then we formalise coalgebras
(Section~\ref{sec:coq-coalg}) and then we put together these notions to
formalise recursive coalgebras and hylomorphisms
(Section~\ref{sec:coq-rec-coalgebras}).

@@

Catamorphisms are constructed so that they structurally deconstruct the
datatype, call themselves recursively, and then compose the result using an
\coq{F}-algebra.
\begin{coqcode}
Definition cata_f (a : Alg F A) : LFix -> A
  := fix f (x : LFix) := match x with
        | LFix_in x => a {shape := shape x; cont := fun e => f (cont x e)}
        end.
\end{coqcode}
It is easy to prove that catamorphisms respect setoid equalities.  In fact, we
define it as a proper morphism:
\begin{coqcode}
cata : forall `{setoid A}, Alg F A ~> LFix ~> A
\end{coqcode}
Finally, we prove that catamorphisms satisfy their universality property (see
Section~\ref{sec:category-theory}):
\begin{coqcode}
Lemma cata_univ `{eA : setoid A} (alg : Alg F A) (f : LFix ~> A)
    : f =e cata alg <-> f =e alg \o fmap f \o l_out.
\end{coqcode}
In other words, if there is any other \coq{f} with the same structural recursive
shape as the catamorphism on the algebra \coq{alg} then it must be equal to that
catamorphism.

>>>>

Suppose, for example, that we want to write a program that sorts a list of
integers and multiplies them by 2 at the same time. In OCaml, we may write this
function directly:
\begin{minted}{ocaml}
let rec sort_times_two = function
  | [] -> []
  | h :: t -> let (l, r) = partition (fun x -> x < h) t in
  sort_times_two l @ (h * 2) :: sort_times_two r
\end{minted}

@@

Given two functors $F, G : \CatC \to \CatD$, a \emph{natural transformation} is
is a family of morphisms $\phi_{X} : FX \to GX$ indexed
by the objects $X \in \Obj(\CatC)$ and such that it is \emph{natural} in $X$,
that is $ G(f) \cdot \phi_{X} = \phi_{Y} \cdot F(f)$. Intuitively, a natural
transformation transforms the structure of a functor into the structure of
another functor without assuming what is the type of data contained in them, for
example, a program $f : \texttt{List }X \to \texttt{Maybe} X$ which returns
nothing if the list is empty and the head of the list otherwise needs not to
know the type of the data.

@@

For example, for a set $A$ we define the functor $F : \Set \to \Set$ mapping
$X \mapsto 1 + A \times X$. An $F$-algebra is a set $B$ together with a
structure map $[base, step] : 1 + A \times B \to B$. The initial $F$-algebra is
clearly the set of lists $\List{A}$ and the catamorphism associated with the
type of lists is the unique arrow which recursively translates the initial
algebra $[\nil,\cons]$ into the algebra $[base, step]$ while turning the operation
$\nil$ into $base$ and $\cons$ into $step$.  In functional programming this is
commonly referred to as $\foldr : (1 \to B) \to (A \times B \to B) \to \List{A} \to B$. We
can in fact set $\foldr\;base\;step = \cata{[base, step]}$.

@@

For example, for a set of states $X$ and a finite set of labels $L$ we can
define a labelled transition system (LTS) on $X$ as a function $X \to BX$
implementing the \emph{transition system} with $BX = L \times X$.  In
particular, for a state $x_{1} \in X$, $c(x_{1})$ returns a pair $(l, x_{2})$
where $l \in L$ is the observable action and $x_{2} \in X$ is the next state.
In this case, we can even set some notation for the transition map:
\[
  x \xrightarrow{l} y \eqdef c_{X}(x) = (l, y)
\]
The category of $B$-coalgebras, denoted $B\Coalg$, is the category where objects
are $B$-coalgebras and morphisms $f : (X, c_{X}) \to (Y, c_{Y})$ are
$B$-coalgebra homomorphisms $f : A \to B$, that is
$ c_{Y} \circ f = F(f) \circ c_{X}$. The terminal object in this category is
called the terminal, or final, $B$-coalgebra. The carrier of this coalgebra
corresponds to the greatest fixed-point for the functor $B$, denoted by
$(\nu B, \OutIso)$ with $\OutIso$ being the final $B$-coalgebra witnessing the
isomorphism and $\OutIso^{\circ} : B\nu B \to \nu B$ being its inverse.

For example, the terminal coalgebra for the functor $BX = A \times X$ is the set
of infinite streams over the set $A$, that is the greatest solution to the domain equation
$\Str{A} \iso A \times \Str{A}$.  The uniqueness property of terminal
$B$-coalgebras states that for any other $B$-coalgebra $(X, c_{X})$ there exists
a unique $B$-coalgebra homomorphism into the terminal coalgebra
$(\nu B, \OutIso)$ which is denoted by $\ana{c_{X}}$ and pronounced
``\emph{anamorphism}'' or ``\emph{unfold}''. We spell out the uniqueness
property:
\begin{equation}
  \label{eq:ana:unique}
  f = \ana{c_{X}} \iff f = \OutOp \circ B(f) \circ c_{X}
\end{equation}
From the uniqueness property we can derive the fusion law for unfolds:
\begin{equation}
  \label{eq:ana:fusion}
  \ana{c_{Y}} \circ f = \ana{c_{X}}
\end{equation}
for all $B$-coalgebra homomorphisms $f : (X, c_{X}) \to (Y, c_{Y})$.

@@

An example of a recursive coalgebra is the \coq{partition} function used in
quicksort, which has the type $\text{List } A \to B(\text{List } A)$, where the
functor $B$ is defined as $BX = X \times A \times X$. This function deconstructs
a list by selecting a pivot element of type $A$ and splitting the remaining
elements into two sublists.

Notice that the initial algebra for lists corresponds to the functor
$F X = 1 + A \times X$ and its associated algebra map (or its inverse, viewed as
a coalgebra), $\InOp$ has the type $\text{List } A \to F(\text{List
} A)$. Importantly, \coq{partition} is not this map, which means that it does
not arise from the standard initial algebra structure, and thus catamorphisms
here cannot be used to define recursive functions.

Nevertheless, since \coq{partition} always produces sublists that are strictly
smaller than the input list, it still supports a well-founded recursion scheme,
ensuring the existence of a unique solution. The uniqueness property of the
hylomorphisms yields the following fusion laws:
\begin{align}
  \label{eq:hylo:fusion}
  f \circ \hylo{c}{a} & = \hylo{c}{a'}
  \hspace{1cm}
  \Longleftarrow
  \hspace{1cm}
  f \circ a = a' \circ F(f) \\
  \hylo{c}{a} \circ f & = \hylo{c'}{a}
  \hspace{1cm}
  \Longleftarrow
  \hspace{1cm}
  c \circ f = F(f) \circ c'
\end{align}
Using the hylomorphism fusion laws, we can prove the well-known \emph{deforestation
optimisation}, also known as the \emph{composition
law}~\cite{DBLP:conf/ifl/HinzeHJ10}. This is when two consecutive recursive
computations, one that builds a data structure, and another one that consumes
it, can be fused together into a single recursive definition. This, in turn,
allows us to prove that a recursive hylomorphism is the composition of a
catamorphism and a recursive anamorphism.

@@

\section{Formalising Recursion Schemes}
\label{sec:recursion-schemes}
%Recursion schemes provide an abstract way to consume and generate data.
We now consider the formalisation of hylomorphisms in Coq. We first formalise
container functors as a tool to represent polynomial functors
(Section~\ref{sec:containers}). Then we formalise algebras for container
extensions (Section~\ref{sec:coq-algebras}), then we formalise coalgebras
(Section~\ref{sec:coq-coalg}) and then we put together these notions to
formalise recursive coalgebras and hylomorphisms
(Section~\ref{sec:coq-rec-coalgebras}).

@@

Catamorphisms are constructed so that they structurally deconstruct the
datatype, call themselves recursively, and then compose the result using an
\coq{F}-algebra.
\begin{coqcode}
Definition cata_f (a : Alg F A) : LFix -> A
  := fix f (x : LFix) := match x with
        | LFix_in x => a {shape := shape x; cont := fun e => f (cont x e)}
        end.
\end{coqcode}
It is easy to prove that catamorphisms respect setoid equalities.  In fact, we
define it as a proper morphism:
\begin{coqcode}
cata : forall `{setoid A}, Alg F A ~> LFix ~> A
\end{coqcode}
Finally, we prove that catamorphisms satisfy their universal property (see
Section~\ref{sec:category-theory}):
\begin{coqcode}
Lemma cata_univ `{eA : setoid A} (alg : Alg F A) (f : LFix ~> A)
    : f =e cata alg <-> f =e alg \o fmap f \o l_out.
\end{coqcode}
In other words, if there is any other \coq{f} with the same structural recursive
shape as the catamorphism on the algebra \coq{alg} then it must be equal to that
catamorphism.
